---
title: "NHANES Data Analysis - Combined Overview"
author: "STAT Capstone Project"
date: "`r Sys.Date()`"
output:
  html_notebook:
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6
)
```

```{r load-packages}
library(haven)
library(dplyr)
library(ggplot2)
```

# Data Loading

```{r load-all-data}
dr1iff <- read_xpt("DR1IFF_L (1).xpt")
paq <- read_xpt("PAQ_L (2).xpt")
dbq <- read_xpt("DBQ_L (1).xpt")

cat("DR1IFF: ", nrow(dr1iff), " rows, ", ncol(dr1iff), " columns\n")
cat("PAQ: ", nrow(paq), " rows, ", ncol(paq), " columns\n")
cat("DBQ: ", nrow(dbq), " rows, ", ncol(dbq), " columns\n")

cat("\nChecking DR1IFF variables...\n")
if("DR1IKCAL" %in% names(dr1iff)) {
  cat("DR1IKCAL found\n")
} else {
  cat("DR1IKCAL NOT found. Available variables:\n")
  print(head(names(dr1iff), 20))
}
```

# Dataset 1: DR1IFF - Dietary Interview Day 1

## Variables Overview

```{r dr1iff-variables}
library(haven)
library(dplyr)
library(tidyr)

dr1iff_data <- read_xpt("DR1IFF_L (1).xpt")

# Key variable name mappings (most important ones)
key_var_names <- data.frame(
  variable = c("SEQN", "WTDRD1", "WTDR2D", "DR1ILINE", "DR1DRSTZ", "DR1EXMER", 
               "DRABF", "DRDINT", "DR1DBIH", "DR1DAY", "DR1LANG", "DR1CCMNM", 
               "DR1CCMTX", "DR1_020", "DR1_030Z", "DR1FS", "DR1_040Z", "DR1IFDCD",
               "DR1IGRMS", "DR1IKCAL", "DR1IPROT", "DR1ICARB", "DR1ISUGR", "DR1IFIBE",
               "DR1ITFAT", "DR1ISFAT", "DR1IMFAT", "DR1IPFAT", "DR1ICHOL", "DR1IATOC",
               "DR1IATOA", "DR1IRET", "DR1IVARA", "DR1IACAR", "DR1IBCAR", "DR1ICRYP",
               "DR1ILYCO", "DR1IVB1", "DR1IVB2", "DR1INIAC", "DR1IVB6", "DR1IFOLA",
               "DR1IVB12", "DR1IVC", "DR1IVD", "DR1IVK", "DR1ICALC", "DR1IPHOS",
               "DR1IMAGN", "DR1IIRON", "DR1IZINC", "DR1ICOPP", "DR1ISODI", "DR1IPOTA",
               "DR1ISELE", "DR1ICAFF", "DR1ITHEO", "DR1IALCO"),
  description = c(
    "Respondent sequence number",
    "Dietary day one sample weight",
    "Dietary two-day sample weight",
    "Food/Individual component number",
    "Dietary recall status",
    "Interviewer ID code",
    "Breast-fed infant (either day)",
    "Number of days of intake",
    "Days between intake and HH interview",
    "Day of week of intake",
    "Language of interview",
    "Combination food number",
    "Combination food type",
    "Time of first eating occasion",
    "Time of first eating occasion (minutes)",
    "Food source",
    "Time of eating occasion (minutes)",
    "Food code",
    "Amount consumed (grams)",
    "Energy (kcal)",
    "Protein (g)",
    "Carbohydrates (g)",
    "Total sugars (g)",
    "Dietary fiber (g)",
    "Total fat (g)",
    "Saturated fatty acids (g)",
    "Monounsaturated fatty acids (g)",
    "Polyunsaturated fatty acids (g)",
    "Cholesterol (mg)",
    "Vitamin E as alpha-tocopherol (mg)",
    "Vitamin E as added alpha-tocopherol (mg)",
    "Retinol (mcg)",
    "Vitamin A, RAE (mcg)",
    "Alpha-carotene (mcg)",
    "Beta-carotene (mcg)",
    "Beta-cryptoxanthin (mcg)",
    "Lycopene (mcg)",
    "Thiamin (Vitamin B1) (mg)",
    "Riboflavin (Vitamin B2) (mg)",
    "Niacin (mg)",
    "Vitamin B6 (mg)",
    "Folate, DFE (mcg)",
    "Vitamin B12 (mcg)",
    "Vitamin C (mg)",
    "Vitamin D (mcg)",
    "Vitamin K (mcg)",
    "Calcium (mg)",
    "Phosphorus (mg)",
    "Magnesium (mg)",
    "Iron (mg)",
    "Zinc (mg)",
    "Copper (mg)",
    "Sodium (mg)",
    "Potassium (mg)",
    "Selenium (mcg)",
    "Caffeine (mg)",
    "Theobromine (mg)",
    "Alcohol (g)"
  )
)

cat("Total variables:", ncol(dr1iff_data), "\n")
cat("Total records:", nrow(dr1iff_data), "\n\n")

# Show all variables with missing data counts
missing_summary <- dr1iff_data %>%
  summarise_all(~ sum(is.na(.))) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "missing_count") %>%
  mutate(
    total_records = nrow(dr1iff_data),
    missing_pct = (missing_count / total_records) * 100,
    non_missing = total_records - missing_count
  ) %>%
  left_join(key_var_names, by = "variable") %>%
  mutate(
    description = ifelse(is.na(description), paste("Variable:", variable), description)
  ) %>%
  select(variable, description, missing_count, non_missing, missing_pct) %>%
  arrange(desc(missing_count))

print(missing_summary)

cat("\nKey variable names:\n")
for(i in 1:min(20, nrow(key_var_names))) {
  cat("  -", key_var_names$variable[i], ":", key_var_names$description[i], "\n")
}
cat("  ... and", ncol(dr1iff_data) - 20, "more variables\n")
```

## Summary Statistics

```{r dr1iff-summary}
library(haven)
library(dplyr)

dr1iff_data <- read_xpt("DR1IFF_L (1).xpt")

cat("Loaded dataset: ", nrow(dr1iff_data), " rows, ", ncol(dr1iff_data), " columns\n")
cat("Has DR1IKCAL? ", "DR1IKCAL" %in% names(dr1iff_data), "\n\n")

if("DR1IKCAL" %in% names(dr1iff_data) && nrow(dr1iff_data) > 100000) {
  dr1iff_summary <- dr1iff_data %>%
    summarise(
      total_records = n(),
      mean_calories = mean(DR1IKCAL, na.rm = TRUE),
      mean_protein = mean(DR1IPROT, na.rm = TRUE),
      mean_carbs = mean(DR1ICARB, na.rm = TRUE),
      mean_fat = mean(DR1ITFAT, na.rm = TRUE),
      mean_sugar = mean(DR1ISUGR, na.rm = TRUE),
      mean_fiber = mean(DR1IFIBE, na.rm = TRUE),
      mean_cholesterol = mean(DR1ICHOL, na.rm = TRUE)
    )
  print(dr1iff_summary)
} else {
  cat("ERROR: Wrong dataset loaded or variables missing!\n")
  cat("Expected: ~100,000 rows, 84 columns\n")
  cat("Got: ", nrow(dr1iff_data), " rows, ", ncol(dr1iff_data), " columns\n")
  cat("First few variable names:\n")
  print(head(names(dr1iff_data), 10))
}
```

## Calories Distribution

```{r dr1iff-calories-plot}
library(haven)
library(ggplot2)
library(dplyr)

dr1iff_data <- read_xpt("DR1IFF_L (1).xpt")

if("DR1IKCAL" %in% names(dr1iff_data) && nrow(dr1iff_data) > 100000) {
  # Aggregate by person to get daily totals
  daily_totals <- dr1iff_data %>%
    group_by(SEQN) %>%
    summarise(
      daily_calories = sum(DR1IKCAL, na.rm = TRUE)
    )
  
  # Create 10 calorie groups with more granularity in middle ranges
  max_cal <- max(daily_totals$daily_calories, na.rm = TRUE)
  
  # Create custom breaks: more splits in middle ranges (where most people are)
  # Smaller bins: 0-1000, 1000-1500, 1500-2000, 2000-2500, 2500-3000
  # Larger bins for higher ranges: 3000-4000, 4000-5000, 5000-6000, 6000-8000, 8000+
  breaks <- c(0, 1000, 1500, 2000, 2500, 3000, 4000, 5000, 6000, 8000, max_cal)
  
  # Create labels with calorie ranges
  group_labels <- paste0(round(breaks[-length(breaks)], 0), "-", round(breaks[-1], 0), " kcal")
  # Fix the last label to show "8000+"
  group_labels[length(group_labels)] <- paste0("8000+ kcal")
  
  daily_totals <- daily_totals %>%
    mutate(
      calorie_group = cut(daily_calories, breaks = breaks, 
                          include.lowest = TRUE, 
                          labels = group_labels)
    )
  
  # Calculate average calories per group and count of people
  group_summary <- daily_totals %>%
    group_by(calorie_group) %>%
    summarise(
      avg_calories = mean(daily_calories, na.rm = TRUE),
      count_people = n(),
      .groups = 'drop'
    )
  
  print(group_summary)
  
  # Bar chart showing count of people in each group
  ggplot(group_summary, aes(x = calorie_group, y = count_people)) +
    geom_bar(stat = "identity", fill = "steelblue", alpha = 0.7) +
    geom_text(aes(label = count_people), vjust = -0.5, size = 3) +
    labs(title = "Number of People in Each Calorie Group (10 Groups)",
         x = "Calorie Range",
         y = "Number of People",
         subtitle = paste("Average calories per group shown in table above")) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
} else {
  cat("ERROR: Wrong dataset or DR1IKCAL variable not available.\n")
  cat("Dataset has", nrow(dr1iff_data), "rows and", ncol(dr1iff_data), "columns\n")
}
```

# Dataset 2: PAQ - Physical Activity Questionnaire

## Variables Overview

```{r paq-variables}
library(haven)
library(dplyr)
library(tidyr)

paq <- read_xpt("PAQ_L (2).xpt")

# Variable name mapping
var_names <- data.frame(
  variable = c("SEQN", "PAD790Q", "PAD790U", "PAD800", "PAD810Q", "PAD810U", "PAD820", "PAD680"),
  description = c(
    "Respondent sequence number",
    "Work activity quantity",
    "Work activity unit",
    "Work activity minutes per week",
    "Transportation activity quantity",
    "Transportation activity unit",
    "Transportation activity minutes per week",
    "Total moderate-intensity work activity minutes per week"
  )
)

cat("Total variables:", ncol(paq), "\n")
cat("Total records:", nrow(paq), "\n\n")

# Show all variables with missing data counts
missing_summary <- paq %>%
  summarise_all(~ sum(is.na(.))) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "missing_count") %>%
  mutate(
    total_records = nrow(paq),
    missing_pct = (missing_count / total_records) * 100,
    non_missing = total_records - missing_count
  ) %>%
  left_join(var_names, by = "variable") %>%
  select(variable, description, missing_count, non_missing, missing_pct) %>%
  arrange(desc(missing_count))

print(missing_summary)

cat("\nAll variable names:\n")
for(i in 1:nrow(var_names)) {
  cat("  -", var_names$variable[i], ":", var_names$description[i], "\n")
}
```

## Summary Statistics

```{r paq-summary}
paq <- read_xpt("PAQ_L (2).xpt")

paq_summary <- paq %>%
  summarise(
    total_records = n(),
    mean_work_minutes = mean(PAD800, na.rm = TRUE),
    mean_transport_minutes = mean(PAD820, na.rm = TRUE),
    mean_moderate_activity = mean(PAD680, na.rm = TRUE)
  )
print(paq_summary)
```

## Physical Activity Distribution

```{r paq-activity-plot}
library(haven)
library(ggplot2)
library(dplyr)

paq <- read_xpt("PAQ_L (2).xpt")

# Filter out missing values and extreme outliers
paq_clean <- paq %>%
  filter(!is.na(PAD680), PAD680 < 5000)

# Create 10 activity groups with more granularity in lower ranges
max_activity <- max(paq_clean$PAD680, na.rm = TRUE)

# Create custom breaks: more splits in lower ranges (where most people are)
# Smaller bins: 0-100, 100-200, 200-300, 300-400, 400-500, 500-600
# Larger bins for higher ranges: 600-800, 800-1000, 1000-1500, 1500+
breaks <- c(0, 100, 200, 300, 400, 500, 600, 800, 1000, 1500, max_activity)

# Create labels with activity ranges
group_labels <- paste0(round(breaks[-length(breaks)], 0), "-", round(breaks[-1], 0), " min/week")
# Fix the last label
group_labels[length(group_labels)] <- paste0("1500+ min/week")

paq_clean <- paq_clean %>%
  mutate(
    activity_group = cut(PAD680, breaks = breaks, 
                         include.lowest = TRUE, 
                         labels = group_labels)
  )

# Calculate average activity per group and count of people
group_summary <- paq_clean %>%
  group_by(activity_group) %>%
  summarise(
    avg_minutes = mean(PAD680, na.rm = TRUE),
    count_people = n(),
    .groups = 'drop'
  )

print(group_summary)

# Bar chart showing count of people in each group
ggplot(group_summary, aes(x = activity_group, y = count_people)) +
  geom_bar(stat = "identity", fill = "darkgreen", alpha = 0.7) +
  geom_text(aes(label = count_people), vjust = -0.5, size = 3) +
  labs(title = "Number of People in Each Activity Group (10 Groups)",
       x = "Activity Range (Minutes per Week)",
       y = "Number of People",
       subtitle = paste("Average minutes per group shown in table above")) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

# Dataset 3: DBQ - Diet Behavior Questionnaire

## Variables Overview

```{r dbq-variables}
library(haven)
library(dplyr)
library(tidyr)

dbq <- read_xpt("DBQ_L (1).xpt")

# Variable name mapping
var_names <- data.frame(
  variable = c("SEQN", "DBQ010", "DBD030", "DBD041", "DBD050", "DBD055", "DBD061",
               "DBQ073A", "DBQ073B", "DBQ073C", "DBQ073D", "DBQ073E", "DBQ073U",
               "DBQ301", "DBQ330", "DBQ360", "DBQ370", "DBD381", "DBQ390", "DBQ400",
               "DBD411", "DBQ421", "DBQ424", "DBQ930", "DBQ935", "DBQ940", "DBQ945"),
  description = c(
    "Respondent sequence number",
    "How healthy is your diet",
    "Days since last ate fish",
    "Days since last ate beans",
    "Days since last ate peanuts",
    "Days since last ate other nuts",
    "Days since last ate seeds",
    "Frequency: meals from fast food/pizza place",
    "Frequency: meals from restaurant with waiter",
    "Frequency: meals from restaurant without waiter",
    "Frequency: meals from store (deli/supermarket)",
    "Frequency: meals from vending machine",
    "Frequency: meals from other source",
    "How often do you eat meals from fast food/pizza place",
    "How often do you eat meals from restaurant with waiter",
    "How often do you eat meals from restaurant without waiter",
    "How often do you eat meals from store (deli/supermarket)",
    "Days since last ate meals from store",
    "How often do you eat meals from vending machine",
    "How often do you eat meals from other source",
    "Days since last ate meals from other source",
    "How often do you eat meals prepared away from home",
    "How often do you eat meals prepared away from home (past 7 days)",
    "How often do you eat meals prepared away from home (past 30 days)",
    "How often do you eat meals prepared away from home (past 12 months)",
    "How often do you eat meals prepared away from home (lifetime)",
    "How often do you eat meals prepared away from home (other)"
  )
)

cat("Total variables:", ncol(dbq), "\n")
cat("Total records:", nrow(dbq), "\n\n")

# Show all variables with missing data counts
missing_summary <- dbq %>%
  summarise_all(~ sum(is.na(.))) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "missing_count") %>%
  mutate(
    total_records = nrow(dbq),
    missing_pct = (missing_count / total_records) * 100,
    non_missing = total_records - missing_count
  ) %>%
  left_join(var_names, by = "variable") %>%
  select(variable, description, missing_count, non_missing, missing_pct) %>%
  arrange(desc(missing_count))

print(missing_summary)

cat("\nAll variable names:\n")
for(i in 1:nrow(var_names)) {
  cat("  -", var_names$variable[i], ":", var_names$description[i], "\n")
}
```

## Summary Statistics

```{r dbq-summary}
dbq <- read_xpt("DBQ_L (1).xpt")

dbq_summary <- dbq %>%
  summarise(
    total_records = n(),
    mean_diet_health = mean(DBQ010, na.rm = TRUE),
    mean_days_since_fish = mean(DBD030, na.rm = TRUE),
    mean_days_since_beans = mean(DBD041, na.rm = TRUE),
    mean_days_since_peanuts = mean(DBD050, na.rm = TRUE)
  )
print(dbq_summary)
```

## Diet Health Rating Distribution

```{r dbq-health-plot}
dbq <- read_xpt("DBQ_L (1).xpt")

dbq_health <- dbq %>%
  filter(!is.na(DBQ010)) %>%
  count(DBQ010) %>%
  mutate(health_label = case_when(
    DBQ010 == 1 ~ "Excellent",
    DBQ010 == 2 ~ "Very Good",
    DBQ010 == 3 ~ "Good",
    DBQ010 == 4 ~ "Fair",
    DBQ010 == 5 ~ "Poor"
  ))

ggplot(dbq_health, aes(x = reorder(health_label, DBQ010), y = n)) +
  geom_bar(stat = "identity", fill = "coral", alpha = 0.7) +
  labs(title = "Distribution of Self-Reported Diet Health",
       x = "Diet Health Rating",
       y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

# Combined Analysis

## Linking Datasets by SEQN

```{r combined-overview}
library(haven)
library(dplyr)

dr1iff <- read_xpt("DR1IFF_L (1).xpt")
paq <- read_xpt("PAQ_L (2).xpt")
dbq <- read_xpt("DBQ_L (1).xpt")

cat("=== ID Variable: SEQN (Respondent Sequence Number) ===\n\n")

# Get unique SEQN from each dataset
dr1iff_seqn <- unique(dr1iff$SEQN)
paq_seqn <- unique(paq$SEQN)
dbq_seqn <- unique(dbq$SEQN)

cat("Unique people in each dataset:\n")
cat("  DR1IFF:", length(dr1iff_seqn), "people\n")
cat("  PAQ:", length(paq_seqn), "people\n")
cat("  DBQ:", length(dbq_seqn), "people\n\n")

# Find overlaps
common_all <- intersect(intersect(dr1iff_seqn, paq_seqn), dbq_seqn)
common_dr1iff_paq <- intersect(dr1iff_seqn, paq_seqn)
common_dr1iff_dbq <- intersect(dr1iff_seqn, dbq_seqn)
common_paq_dbq <- intersect(paq_seqn, dbq_seqn)

cat("Overlap between datasets:\n")
cat("  All 3 datasets:", length(common_all), "people\n")
cat("  DR1IFF + PAQ:", length(common_dr1iff_paq), "people\n")
cat("  DR1IFF + DBQ:", length(common_dr1iff_dbq), "people\n")
cat("  PAQ + DBQ:", length(common_paq_dbq), "people\n\n")

cat("Coverage:\n")
cat("  DR1IFF people also in DBQ:", round(length(common_dr1iff_dbq)/length(dr1iff_seqn)*100, 1), "%\n")
cat("  PAQ people also in DBQ:", round(length(common_paq_dbq)/length(paq_seqn)*100, 1), "%\n")
cat("  People in all 3 datasets:", round(length(common_all)/length(dbq_seqn)*100, 1), "% of DBQ\n\n")

if(length(common_all) > 0) {
  cat("Sample of common SEQN (people in all 3 datasets):\n")
  cat("  ", paste(head(common_all, 10), collapse = ", "), "\n")
  cat("\n✓ SEQN can be used to link/merge all three datasets!\n")
  cat("  You can merge on SEQN to combine dietary, activity, and behavior data.\n")
}
```

## Merging and Exploring Linked Data

```{r merge-datasets}
library(haven)
library(dplyr)

dr1iff_data <- read_xpt("DR1IFF_L (1).xpt")
paq <- read_xpt("PAQ_L (2).xpt")
dbq <- read_xpt("DBQ_L (1).xpt")

dr1iff_subset <- dr1iff_data %>%
  select(SEQN, DR1IKCAL, DR1IPROT, DR1ICARB, DR1ITFAT, DR1ISUGR, DR1IFIBE, DR1ICHOL)

dr1iff_person <- dr1iff_subset %>%
  group_by(SEQN) %>%
  summarise(
    total_calories = sum(DR1IKCAL, na.rm = TRUE),
    total_protein = sum(DR1IPROT, na.rm = TRUE),
    total_carbs = sum(DR1ICARB, na.rm = TRUE),
    total_fat = sum(DR1ITFAT, na.rm = TRUE),
    total_sugar = sum(DR1ISUGR, na.rm = TRUE),
    total_fiber = sum(DR1IFIBE, na.rm = TRUE),
    total_cholesterol = sum(DR1ICHOL, na.rm = TRUE),
    .groups = 'drop'
  )

merged_data <- dr1iff_person %>%
  left_join(paq, by = "SEQN") %>%
  left_join(dbq, by = "SEQN")

cat("Combined dataset: ", nrow(merged_data), " people, ", ncol(merged_data), " variables\n")

merged_data
```

## Combined Dataset Exploration

```{r combined-exploration}
library(dplyr)
library(tidyr)

# Variable name mapping for all variables in merged dataset
var_descriptions <- data.frame(
  variable = c("SEQN", "total_calories", "total_protein", "total_carbs", "total_fat", 
               "total_sugar", "total_fiber", "total_cholesterol", "PAD790Q", "PAD790U",
               "PAD800", "PAD810Q", "PAD810U", "PAD820", "PAD680", "DBQ010", "DBD030",
               "DBD041", "DBD050", "DBD055", "DBD061", "DBQ073A", "DBQ073B", "DBQ073C",
               "DBQ073D", "DBQ073E", "DBQ073U", "DBQ301", "DBQ330", "DBQ360", "DBQ370",
               "DBD381", "DBQ390", "DBQ400", "DBD411", "DBQ421", "DBQ424", "DBQ930",
               "DBQ935", "DBQ940", "DBQ945"),
  description = c(
    "Respondent sequence number",
    "Total calories per day (kcal)",
    "Total protein per day (g)",
    "Total carbohydrates per day (g)",
    "Total fat per day (g)",
    "Total sugar per day (g)",
    "Total dietary fiber per day (g)",
    "Total cholesterol per day (mg)",
    "Work activity quantity",
    "Work activity unit",
    "Work activity minutes per week",
    "Transportation activity quantity",
    "Transportation activity unit",
    "Transportation activity minutes per week",
    "Total moderate-intensity activity minutes per week",
    "How healthy is your diet",
    "Days since last ate fish",
    "Days since last ate beans",
    "Days since last ate peanuts",
    "Days since last ate other nuts",
    "Days since last ate seeds",
    "Frequency: meals from fast food/pizza place",
    "Frequency: meals from restaurant with waiter",
    "Frequency: meals from restaurant without waiter",
    "Frequency: meals from store (deli/supermarket)",
    "Frequency: meals from vending machine",
    "Frequency: meals from other source",
    "How often: meals from fast food/pizza place",
    "How often: meals from restaurant with waiter",
    "How often: meals from restaurant without waiter",
    "How often: meals from store (deli/supermarket)",
    "Days since last ate meals from store",
    "How often: meals from vending machine",
    "How often: meals from other source",
    "Days since last ate meals from other source",
    "How often: meals prepared away from home",
    "How often: meals prepared away from home (past 7 days)",
    "How often: meals prepared away from home (past 30 days)",
    "How often: meals prepared away from home (past 12 months)",
    "How often: meals prepared away from home (lifetime)",
    "How often: meals prepared away from home (other)"
  )
)

cat("=== COMBINED DATASET OVERVIEW ===\n\n")

cat("1. Dataset Dimensions:\n")
cat("   Rows (people):", nrow(merged_data), "\n")
cat("   Columns (variables):", ncol(merged_data), "\n")
cat("   Total data points:", nrow(merged_data) * ncol(merged_data), "\n\n")

cat("2. All Variable Names:\n")
var_names_with_desc <- var_descriptions %>%
  filter(variable %in% names(merged_data)) %>%
  mutate(full_name = paste0(variable, " - ", description))
print(var_names_with_desc %>% select(variable, description))
cat("\n")

cat("3. Missing Values Summary:\n")
missing_summary <- merged_data %>%
  summarise_all(~ sum(is.na(.))) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "missing_count") %>%
  mutate(
    total_obs = nrow(merged_data),
    missing_pct = (missing_count / total_obs) * 100,
    non_missing = total_obs - missing_count
  ) %>%
  left_join(var_descriptions, by = "variable") %>%
  mutate(
    description = ifelse(is.na(description), paste("Variable:", variable), description)
  ) %>%
  select(variable, description, missing_count, non_missing, missing_pct) %>%
  arrange(desc(missing_count))

print(missing_summary)
cat("\n")

cat("4. Total Missing Values:\n")
cat("   Total missing:", sum(missing_summary$missing_count), "\n")
cat("   Percentage missing:", round(sum(missing_summary$missing_count) / (nrow(merged_data) * ncol(merged_data)) * 100, 2), "%\n\n")

cat("5. Variables with No Missing Data:\n")
complete_vars <- missing_summary %>%
  filter(missing_count == 0) %>%
  select(variable, description)
print(complete_vars)
cat("   Count:", nrow(complete_vars), "variables\n\n")

cat("6. Variables with Most Missing Data (Top 10):\n")
print(head(missing_summary %>% select(variable, description, missing_count, missing_pct), 10))
cat("\n")

cat("7. Data Structure:\n")
str(merged_data, give.attr = FALSE, list.len = 10)
cat("\n")

cat("8. Summary Statistics (Key Variables):\n")
if("total_calories" %in% names(merged_data)) {
  key_summary <- merged_data %>%
    select(SEQN, total_calories, total_protein, total_fat, PAD680, DBQ010) %>%
    summarise(
      people_with_calories = sum(!is.na(total_calories)),
      people_with_activity = sum(!is.na(PAD680)),
      people_with_diet_health = sum(!is.na(DBQ010)),
      mean_calories = mean(total_calories, na.rm = TRUE),
      mean_activity = mean(PAD680, na.rm = TRUE),
      mean_diet_health = mean(DBQ010, na.rm = TRUE)
    )
  print(key_summary)
}
cat("\n")

cat("9. Sample of Combined Data (First 10 rows, key columns):\n")
sample_cols <- c("SEQN", "total_calories", "total_protein", "PAD680", "DBQ010")
available_cols <- sample_cols[sample_cols %in% names(merged_data)]
if(length(available_cols) > 0) {
  print(head(merged_data %>% select(all_of(available_cols)), 10))
}
cat("\n")

cat("10. People with Data in Multiple Datasets:\n")
overlap_summary <- merged_data %>%
  summarise(
    total_people = n(),
    has_dietary = sum(!is.na(total_calories)),
    has_activity = sum(!is.na(PAD680)),
    has_behavior = sum(!is.na(DBQ010)),
    has_dietary_and_activity = sum(!is.na(total_calories) & !is.na(PAD680)),
    has_dietary_and_behavior = sum(!is.na(total_calories) & !is.na(DBQ010)),
    has_activity_and_behavior = sum(!is.na(PAD680) & !is.na(DBQ010)),
    has_all_three = sum(!is.na(total_calories) & !is.na(PAD680) & !is.na(DBQ010))
  )
print(overlap_summary)
cat("\n")

cat("11. Data Completeness by Person:\n")
completeness <- merged_data %>%
  mutate(
    total_vars = ncol(merged_data),
    missing_count = rowSums(is.na(select(., -SEQN))),
    complete_count = total_vars - missing_count - 1,
    completeness_pct = (complete_count / (total_vars - 1)) * 100
  ) %>%
  select(SEQN, missing_count, complete_count, completeness_pct)

completeness_summary <- completeness %>%
  summarise(
    people_with_no_missing = sum(missing_count == 0),
    people_with_90pct_complete = sum(completeness_pct >= 90),
    people_with_80pct_complete = sum(completeness_pct >= 80),
    people_with_70pct_complete = sum(completeness_pct >= 70),
    people_with_50pct_complete = sum(completeness_pct >= 50),
    mean_completeness = mean(completeness_pct),
    median_completeness = median(completeness_pct)
  )
print(completeness_summary)
cat("\n")

cat("Distribution of Completeness:\n")
completeness_dist <- completeness %>%
  mutate(completeness_group = case_when(
    completeness_pct == 100 ~ "100% (Complete)",
    completeness_pct >= 90 ~ "90-99%",
    completeness_pct >= 80 ~ "80-89%",
    completeness_pct >= 70 ~ "70-79%",
    completeness_pct >= 50 ~ "50-69%",
    TRUE ~ "<50%"
  )) %>%
  count(completeness_group) %>%
  arrange(desc(completeness_group))
print(completeness_dist)
cat("\n")

cat("Sample of most complete records (top 10):\n")
print(head(completeness %>% arrange(desc(completeness_pct)), 10))
```

## Completeness: DR1IFF + PAQ Only (Excluding DBQ)

```{r completeness-diet-activity}
library(haven)
library(dplyr)

dr1iff_data <- read_xpt("DR1IFF_L (1).xpt")
paq <- read_xpt("PAQ_L (2).xpt")

dr1iff_subset <- dr1iff_data %>%
  select(SEQN, DR1IKCAL, DR1IPROT, DR1ICARB, DR1ITFAT, DR1ISUGR, DR1IFIBE, DR1ICHOL)

dr1iff_person <- dr1iff_subset %>%
  group_by(SEQN) %>%
  summarise(
    total_calories = sum(DR1IKCAL, na.rm = TRUE),
    total_protein = sum(DR1IPROT, na.rm = TRUE),
    total_carbs = sum(DR1ICARB, na.rm = TRUE),
    total_fat = sum(DR1ITFAT, na.rm = TRUE),
    total_sugar = sum(DR1ISUGR, na.rm = TRUE),
    total_fiber = sum(DR1IFIBE, na.rm = TRUE),
    total_cholesterol = sum(DR1ICHOL, na.rm = TRUE),
    .groups = 'drop'
  )

merged_diet_activity <- dr1iff_person %>%
  left_join(paq, by = "SEQN")

cat("=== DR1IFF + PAQ Only (No DBQ) ===\n\n")
cat("Total people:", nrow(merged_diet_activity), "\n")
cat("Total variables:", ncol(merged_diet_activity), "\n\n")

completeness_da <- merged_diet_activity %>%
  mutate(
    total_vars = ncol(merged_diet_activity),
    missing_count = rowSums(is.na(select(., -SEQN))),
    complete_count = total_vars - missing_count - 1,
    completeness_pct = (complete_count / (total_vars - 1)) * 100
  ) %>%
  select(SEQN, missing_count, complete_count, completeness_pct)

completeness_summary_da <- completeness_da %>%
  summarise(
    people_with_no_missing = sum(missing_count == 0),
    people_with_90pct_complete = sum(completeness_pct >= 90),
    people_with_80pct_complete = sum(completeness_pct >= 80),
    people_with_70pct_complete = sum(completeness_pct >= 70),
    mean_completeness = mean(completeness_pct),
    median_completeness = median(completeness_pct)
  )
print(completeness_summary_da)
cat("\n")

cat("Distribution of Completeness:\n")
completeness_dist_da <- completeness_da %>%
  mutate(completeness_group = case_when(
    completeness_pct == 100 ~ "100% (Complete)",
    completeness_pct >= 90 ~ "90-99%",
    completeness_pct >= 80 ~ "80-89%",
    completeness_pct >= 70 ~ "70-79%",
    completeness_pct >= 50 ~ "50-69%",
    TRUE ~ "<50%"
  )) %>%
  count(completeness_group) %>%
  arrange(desc(completeness_group))
print(completeness_dist_da)
cat("\n")

cat("Top 10 most complete records:\n")
print(head(completeness_da %>% arrange(desc(completeness_pct)), 10))
```

## Quick Totals Summary

```{r totals-summary}
dr1iff <- read_xpt("DR1IFF_L (1).xpt")
paq <- read_xpt("PAQ_L (2).xpt")
dbq <- read_xpt("DBQ_L (1).xpt")

cat("=== TOTALS SUMMARY ===\n\n")
cat("DR1IFF Dataset:\n")
cat("  Total records:", nrow(dr1iff), "\n")
if("DR1IKCAL" %in% names(dr1iff)) {
  cat("  Total calories (sum):", sum(dr1iff$DR1IKCAL, na.rm = TRUE), "kcal\n")
  cat("  Total protein (sum):", sum(dr1iff$DR1IPROT, na.rm = TRUE), "g\n")
  cat("  Total fat (sum):", sum(dr1iff$DR1ITFAT, na.rm = TRUE), "g\n")
} else {
  cat("  Nutritional variables not available\n")
}
cat("\n")

cat("PAQ Dataset:\n")
cat("  Total records:", nrow(paq), "\n")
cat("  Total work activity minutes:", sum(paq$PAD800, na.rm = TRUE), "minutes\n")
cat("  Total transport activity minutes:", sum(paq$PAD820, na.rm = TRUE), "minutes\n\n")

cat("DBQ Dataset:\n")
cat("  Total records:", nrow(dbq), "\n")
cat("  Respondents with diet health rating:", sum(!is.na(dbq$DBQ010)), "\n")
```

