---
title: "STAT Capstone Report"
output:
  pdf_document: default
  html_document: default
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Overview

This report runs the modular R analysis pipeline in **numeric order** (`00` → `05`), plus optional `07` poster outputs.  
Each step below calls `source()` on one script file, which executes **every top-level statement** in that file in order (package loads, `ROOT` / paths, function definitions, and any other non-function code).  
Code **inside** functions then runs when the later chunks call `run_*()` (same as running the scripts interactively line-by-line up to each function, then invoking those functions).

**Knit this file with the working directory set to** `macro_goal_app/backend/Rcode` (RStudio: *Knit* from the Rmd’s folder, or `setwd()` before `rmarkdown::render()`).

---

## Pipeline: source `00`–`05` (full file execution each)

```{r rcode-00-project-setup}
source("00_project_setup.R")
```

```{r rcode-01-data-prep}
source("01_data_prep.R")
```

```{r rcode-02-audit-vif-main}
source("02_audit_vif_main.R")
```

```{r rcode-03-interaction-models}
source("03_interaction_models.R")
```

```{r rcode-04-body-composition}
source("04_body_composition_models.R")
```

```{r rcode-05-race-income}
source("05_race_income_models_modular.R")
```

```{r build-data-and-design}
data <- prepare_data()
design <- make_design(data)
```

---

## 1) Missingness, VIF, and Main Effects

```{r main-effects}
main_res <- run_audit_vif_main(data, design, make_plots = TRUE)
source("07_poster_tables_plots.R")
```

## 2) Interaction Models

```{r interaction-models}
int_res <- run_interaction_models(data, design)
```

## 3) Body Composition Models

```{r body-composition}
body_res <- run_body_composition_models(data)
```

## 4) Race + Income Adjusted Models

```{r race-income}
race_income_res <- run_race_income_models(data, design)
```

## Notes

- Core analysis scripts are kept as modular `.R` files for reuse and debugging.
- This `.Rmd` is the presentation/report layer: it **sources** `00`–`05` sequentially, then calls the main entry-point functions above. After main effects, `07_poster_tables_plots.R` writes one-page tables and plots to `Rcode/poster_output/`.
- `01_data_prep.R` begins by sourcing `00_project_setup.R` again (so `00` runs twice when you knit); that is harmless and keeps `01` usable on its own.
- The live API uses `recommend_macros.R`, which sources a slightly different set (includes `06_calorie_models.R`, omits `05`); see that script for deployment behavior.
